---
title: "V506 Lab 4"
author: "Luis Navarro"
date: "Spring 2024"
output:
  slidy_presentation: default
  #ioslides_presentation: default
subtitle: Data Summaries and Functions
---

```{r setup, include=FALSE} 
knitr::opts_chunk$set(warning = FALSE, message = FALSE) 
```

```{r, echo=FALSE, eval=TRUE}
# Clean the environment 
rm(list=ls())
setwd("/Users/luisenriquenavarro/Library/CloudStorage/OneDrive-IndianaUniversity/V506/Lab4")
library(pacman)
p_load(tidyverse, rmarkdown, dplyr, ggplot2, descriptr)
```

# Functions in R

- We have explored several functions from base R and from specific packages. 

- We can create our own functions too! 

- Function structure: relationship between *inputs* and *outputs* through a rule of correspondence (program). 


- Recall math intuition. Let $y$ be a function $f$ of $x$. 

\begin{equation*}
y = f(x)
\end{equation*}

- Function (program) $f$ transforms input $x$ on output $y$. 

- Function syntax in R

```{r, eval = FALSE, echo = TRUE}
functionName = function(inputs) {
< function body >
return(value)
}
```


# Functions: Example 

- Example: Linear Function: 

\begin{equation*}
y = 1 + 2x
\end{equation*}

- In words: take $x$, multiply it by 2, and add 1. (Note: order of operations matter)

```{r, eval = TRUE, echo = TRUE}
line_function <- function(x){
  y = 1 + 2*x
  return(y)
}
```

- Parameters (Input): x. 

- Output: y 


```{r}
x1 <- 5 
y1 <- line_function(x1)
print(c(x1,y1))
```

- You can apply some functions to vectors. 

```{r}
seq_x <- c(1:100)
seq_y <- line_function(seq_x)

# Show data in data frame 
data_line <- data.frame(x = seq_x, y = seq_y) 
data_line %>% head(n=10)
```
```{r}
ggplot(data = data_line, mapping = aes(x = x,y = y)) + 
       geom_line(col="black") + 
  theme_bw()
```

# Function 

- Functions can have more than one parameters.

- Recall the general form of the linear function: $b$ is the intercept, $m$ is the slope, and $x$ is the data. 

\begin{equation*}
y = b + mx
\end{equation*}

```{r, eval = TRUE, echo = TRUE}
line_function1 <- function(x,b,m){
  y = b + m*x
  return(y)
}
```


- **Example:** let $b=10$ and $m=-2$. 

```{r, eval = TRUE, echo = TRUE}
data_line2 <- data.frame(x = seq_x, 
                         y = line_function1(x = seq_x, b = 10, m = -2))

data_line2 %>% head(n=5L) 
```

```{r, eval = TRUE, echo = TRUE}
ggplot(data = data_line2, mapping = aes(x = x,y = y)) + 
       geom_line(col="red")+ theme_bw()
```

# Example: Fast Food Data 

- Let's load the fast food data. It contains observations on the nutritional content of several items from fast food restaurants in the US. 

- Task: build a table of descriptive statistics (mean, median, standard deviation, number of observations, % missing) for a given variable in the data set. 

```{r, eval = TRUE, echo = TRUE}
fastfood <- read.csv(file = "fastfood.csv", header = TRUE) %>% 
  tibble() 
# Show data: Slice_sample takes a random sample of size n 
fastfood %>% 
  slice_sample(n = 5) 
```


# Functions to compute analysis

- You can use functions to perform more complex tasks than drawing lines. 

- Let's create a function that summarizes a variable from the data set. 

```{r, eval = TRUE, echo = TRUE}
descriptive_table <- function(x){
  # Compute the Descriptive Statistics using base R functions
  x_count <- length(x)
  x_miss <- sum(is.na(x))/length(x)
  x_mean <- mean(x, na.rm = TRUE)
  x_median <- median(x, na.rm = TRUE)
  x_sd <- sd(x, na.rm = TRUE)
  table <- data.frame(obs = x_count, 
                      miss = x_miss, mean = x_mean, 
                      median = x_median, sd = x_sd) %>% 
    tibble() 

  # Return Table 
  return(table)
}
```

- Let's try our function. 

```{r, eval = TRUE, echo = TRUE}
calories <- fastfood$calories
descriptive_table(calories)
```


# Apply Functions to Multiple Objects 

- Most of the times you build a function because you will use that program (recipe) several times. 

- We can use `lapply` to apply a function to list of objects. 

- Remember: lists are flexible and can store objects of any class. 

- Tip: to access elements of a list we need to use double brackets notation.

```{r}
# Example: 
calories <- fastfood$calories
carbs <- fastfood$total_carb
total_fat <- fastfood$total_fat
# Create a list of Variables 
variables <- list(calories, carbs, total_fat)
```

- Show the first element of the list 

```{r, echo = TRUE, eval = FALSE}
# Access first element of the list 
variables[[1]]
```

- Apply the function we created to all the elements of the list. 

- Note: the output of lapply is also a list. 

- lapply syntax: lapply(list, function)

```{r, echo = TRUE, eval = TRUE}
variables_descriptive_tables <- lapply(variables, descriptive_table)
```

- Examine the object. Note each element is a data frame. 


```{r}
bind_rows(variables_descriptive_tables[[1]] %>% mutate(varname = "Calories"), 
          variables_descriptive_tables[[2]] %>% mutate(varname = "Carbs"),
          variables_descriptive_tables[[3]] %>% mutate(varname = "Total Fat")) %>% 
  # Relocate allows you to change the order of the variables in the data frame 
  relocate(varname)
```




# summarize() and group_by()
- Today we explore two powerful `dplyr` functions for data analysis: `group_by` and `summarize`

-   group_by() allows you to specify a categorical variable whose categories will create clusters or subsets of the data. Grouping by category.  

-   summarize() creates a new data frame that uses the grouping variables to summarize the observations in the group. 

- Useful summarizing functions, according to the dplyr documentation:

    - Center: mean(), median()
    - Spread: sd(), IQR(), mad()
    - Range: min(), max(), quantile()
    - Position: first(), last(), nth(),
    - Count: n(), n_distinct()
    - Logical: any(), all()


# Example: Data Summary using summarize 

- Use summarize to build the descriptive statistics of variable: `calories`

```{r, eval=TRUE, echo=TRUE}
descriptive_stats_calories <- fastfood %>% 
  summarize(Mean = mean(calories, na.rm = TRUE), 
            SD = sd(calories, na.rm = TRUE),
            Min = min(calories, na.rm = TRUE), 
            Percent25 = quantile(calories, probs = 0.25), 
            Median = quantile(calories, probs = 0.50), 
            Percent75 = quantile(calories, probs = 0.75), 
            Max = max(calories, na.rm = TRUE), 
            Obs = n()) 
# Show the Table
descriptive_stats_calories 
```
- Tip: Add the variable name to display the table better. 

```{r, eval=TRUE, echo=TRUE}
descriptive_stats_calories %>% 
  mutate(Variable = "Calories") %>% 
  # Function Relocate changes the ordering of the variables 
  relocate(Variable, Obs, Mean, SD)
```

# Example: Data Summary by Category

- Suppose you want to compute the descriptive statistics of the variable calories across restaurants. That is, compare their distributions. 

- We can do that by adding `group_by` to our previous code. 

```{r, eval=TRUE, echo=TRUE}
descriptive_stats_calories_restaurant <- fastfood %>% 
  # Add Grouping Criteria 
  group_by(restaurant) %>% 
  # Same code as before 
  summarize(Mean = mean(calories, na.rm = TRUE), 
            SD = sd(calories, na.rm = TRUE),
            Min = min(calories, na.rm = TRUE), 
            Percent25 = quantile(calories, probs = 0.25), 
            Median = quantile(calories, probs = 0.50), 
            Percent75 = quantile(calories, probs = 0.75), 
            Max = max(calories, na.rm = TRUE), 
            Obs = n()) 
# Show the Table
descriptive_stats_calories_restaurant %>% 
  # Sort the data by highest average calories 
  arrange(-Mean)
```



# Extra Example: Data Summary Across Variables 

- You can do similar analysis and look at descriptive statistics of different variables 

```{r, eval=TRUE, echo=TRUE}
descriptive_stats_calories_restaurant <- fastfood %>% 
  # Add Grouping Criteria 
  group_by(restaurant) %>% 
  # Same code as before 
  summarize(mean_calories = mean(calories, na.rm = TRUE), 
            mean_total_fat = mean(total_fat, na.rm = TRUE), 
            mean_trans_fat = mean(trans_fat, na.rm = TRUE), 
            mean_total_carbs = mean(total_carb, na.rm = TRUE),
            Obs = n()) 
# Show 
descriptive_stats_calories_restaurant %>% 
  arrange(-mean_calories)
```


# Additional functions 

**descriptr's ds_freq_table()**

- This returns a frequency table for categorical and continuous data, returning the frequency, cumulative, frequency, frequency percent, and cumulative frequency percent.

```{r, echo = TRUE, eval = TRUE}
if(!require("descriptr")) {
  install.packages("descriptr")
  library(descriptr)
}
```

```{r}
ds_freq_table(fastfood, calories)
```

```{r}
ds_freq_table(fastfood, calories, bins = 10)
```


