---
title: "V506 Lab 1"
output: slidy_presentation
author: Luis Navarro and Dario Salcedo (based on Mallory Dreyer's slides)
date: Spring 2024 
---

# Welcome to V506!

- The purpose of these labs is to learn how to use R and R studio for statistical analysis. 

- **Ground Rules**: 75 minutes labs. i) lecture (40-45 minutes); ii) in-class exercise (15 minutes); iii) Homework Q&A (with TAs) (15-20 minutes). 

- Interactive settings. Ask as many questions as possible. Try running the code on your own computer in class. Verify how it works. 

- Only 6 labs. **ALL REQUIRED**

# Why R? 

- Fast and Free. 

- Open source: R uses packages (written by users). 

- Programming Language + Statistical Analysis (we'll use packages mostly focused on the latter)

- Productivity improvement in the work place. 


# Pipeline for today 

- Intro to R and R studio 

- Intro to R Markdown 

- R packages 

- Working with R: object types 

- Intro to data manipulation 


# Installing R 

- You should have downloaded R and Rstudio before arriving to lab. 

- For this class, you need both R and Rstudio. 

- R is the programming language and the console runs the code. 

- R studio is the user interface. 


# R Markdown 

- R Markdown allows you to create notebooks that integrate code into a document (pdf, slides, html, MS word). 

- File extension: .Rmd

- Excellent tool to present analysis in an organized and clean way. 

- **You can use R Markdown to turn in the assignments** 

```{r, eval = FALSE, echo = TRUE}
install.packages("rmarkdown")
library(rmarkdown)
```

# R Markdown Syntax 

- With markdown you can add snippets of code to a document. 

- To insert a line of code you need to open 

```{r, eval = FALSE, echo = TRUE}
'```{r} 
"Here you write the code"
``` '
```

- You can tell R whether it should evaluate (eval) and/or show (echo) the code in the document.

```{r, eval = FALSE, echo = TRUE}
'```{r, eval = TRUE, echo = TRUE} ``` '
```

# Introduction to Objects in R 

- R is an object oriented programming language. 

- Objects from different types: numbers, strings, atomic vectors, matrices, data frames, tibbles, functions, etc. 

- Each object is assigned a name. 

- '<-' is the assignment operator. 

- For most cases you can also use "=" to assign a name to an object. 

```{r, eval=TRUE, echo=TRUE}
a = 10 
b <- 5
a + b 
```

# R objects and data types 

Let's explore some object types 

```{r, eval=TRUE, echo=TRUE}
a <- 5L 
b <- "Welcome to V506"
c <- TRUE 
d <- 3.14159
e <- c(1,2,3,4)
```

The function class or typeof gives you the object type. 

```{r, eval=TRUE, echo=TRUE}
class(a)
typeof(a)
```

```{r, eval=TRUE, echo=TRUE}
class(a)
class(b)
class(c)
class(d)
```

# Quick review: functions 

- **Definition**: A function is a set of statements organized together to perform a specific task. 

- It is an input-output relationship. 

- Remember from math: $y = f(x)$. For each value of $x$, there is one value of $y$ where $y$ equals to some transformation $f$ of $x$. 

- R has several built-in functions and many more available at user-written packages. 

- You can write your own functions too (more on this later!)

- Example: what does the *print* function do? 

- It takes an object $x$ as input and prints back a type $y$ as output. 

```{r, eval=TRUE, echo=TRUE}
text <- "Hello World"
print(text)
```

# Packages in R

- **Packages:** are a collection of R functions, compiled code and (sometimes) sample data. They are stored in R in *libraries*

- Packages are downloaded from R repositories. Each package has some companionship documentation explaining the functions. Best source of information when debugging your code. 

- To install packages in R you can to look for them at CRAN repository (install packages button) or just use the *install.packages* function. 

- Once installed, you just need to call it using the *library* function. **Notice the syntax** 

```{r, eval = FALSE, echo = TRUE}
install.packages("rmarkdown")
library(rmarkdown)
```

# Data Structures 

- Data sets are collections of observations of one or more variables.  

- **Vectors**: contain one or more elements of the same data type. 

```{r, eval=TRUE, echo=TRUE}
vector1 <- c(1,2,3,4)
print(vector1)
```

- **Matrices**: a table with elements of the same data type organized in rows and columns. 

```{r, eval=TRUE, echo=TRUE}
matrix1 <- matrix(data = vector1, nrow = 2)
print(matrix1)
```

- **Arrays**: n-dimensional collection of data (could have more rows than columns)

```{r, eval=TRUE, echo=TRUE}
array1 <- array(data = c(12:15, 6:10, 3:5), dim = c(2,2,3))
print(array1)
```

- **Lists**: contain one or more elements from potentially different data types. 

```{r, eval=TRUE, echo=TRUE}
list1 <- list(vector1, matrix1, array1,  "Hello World")
print(list1)
```

# Data Structures 

- **Factors**: stores categorical information and assigns levels. 

```{r, eval=TRUE, echo=TRUE}
# Data on the ethnicity of 6 people
eth_vec <- c("white", "white", "hispanic", "black", "asian", "black")
ethnicities <- factor(eth_vec)
ethnicities
```

- **Data Frames**: tabular, can contain different types of data and any number of rows and columns. 

```{r, eval=TRUE, echo=TRUE}
# Data on people
names <- c("Abby", "Bob", "Cady", "Dany", "Emma", "George")
wages <- c(10,8,5,11,6,9)
gender <- factor(c("female","male", "female", "male", "female","male"))
# Create a data frame 
df <- data.frame(names, wages, gender, ethnicities)
print(df)
```
# Intro to Tidyverse and Tibbles 

- Tidyverse is useful package to perform analysis on data frames. We are going to use this a lot. 

```{r, eval=TRUE, echo=TRUE, message = FALSE, warning = FALSE, error = FALSE}
library(tidyverse)
``` 

- **Tibbles**: are data frames, but with some tweaks that allow easier interactions with some functions.   

```{r, eval=TRUE, echo=TRUE}
tibble1 <- tibble(df)
tibble1
``` 
- You can explore the variables of a data frame 

```{r, eval=TRUE, echo=TRUE}
names(tibble1)
``` 

- **Retrieving a variable from a data frame:** To access only one variable of the data frame, just write the name of the data frame followed by a dollar sign "$" and the name of the variable. 

- For example, let's get the vector containing the variable *wages*. 

```{r, eval=TRUE, echo=TRUE}
tibble1$wages
``` 

- Explore its type 

```{r, eval=TRUE, echo=TRUE}
class(tibble1$gender)
```

- Explore some basic descriptive statistics (through the summmary function)

```{r, eval=TRUE, echo=TRUE}
summary(tibble1)
```

- **Create a new variable:** a simple way to add a new variable to the existing data frame is to use the dollar sign "$" syntax. First specify the data frame and then the variable name. 

- Example: create a variable with wages squared. 

```{r, eval=TRUE, echo=TRUE}
tibble1$wages2 = (tibble1$wages)^2
```

- Verify that the data frame updated the number of variables. 

```{r, eval=TRUE, echo=TRUE}
head(tibble1)
```


# Help in R 

- **Tip:** Always check the documentation of the function to see the parameters (inputs) it requires. You can check it by writing a question mark "?" before the name of the function. 

```{r, eval=FALSE, echo=TRUE}
# Help with the print function
?print
```

- Stackoverflow and even ChatGPT help understanding when R gives you an error. 

- *Always read the error message and try to understand it*

- Avoid relying on AI tools for code generation. Use them responsibly. 

# Working in R: some best practices

- When working in projects, try to keep your environment organized. Here are some useful tips. 

- **Define your working directory:** tell R in which folder on your computer you are storing your project (data inputs, code, outputs). 

```{r, eval=TRUE, echo=TRUE}
# Set up the working directory
setwd("C:/Users/luise/OneDrive - Indiana University/v506/Fall23_Labs")
getwd()
```

- **Clean your working environment before beginning to work :** sometimes R will save your past sessions. This could (inadvertently) mix some projects. This code cleans all the objects at the general environment. 

```{r, eval=FALSE, echo=TRUE}
# Clean the working environment
rm(list=ls())
```

- **Always comment your code:** explain what your code is doing at each line. Sometimes you'll revisit a script and won't remember what were you thinking at that time. Furthermore, helps others reviewing your code understanding what's the script doing. With "#" you can comment stuff. 

```{r, eval=TRUE, echo=TRUE}
#Define two integers 
a = 5L
b = 6L 
# Add them up 
c = a+b
#Show the result 
print(c)
```

# Glimpse of data manipulation. 

- For data manipulation and wrangling, we will use the *dplyr* library. 

```{r, eval=TRUE, echo=TRUE, message = FALSE, warning = FALSE, error = FALSE}
install.packages("dplyr")
library(dplyr)
```

- A common task when manipulating data is creating subsets of the data set. 

- For that task, we can use the *filter* function. 

- Example: let's look at the females in our example data frame. 

```{r, eval=TRUE, echo=TRUE}
filter(tibble1, gender == "female")
```

- More on this next lab! 

# Loading data into R 

- For a real task, the first thing to do is load your data set into R. 

- There are functions specific for each data type (csv, xlsx, dta, html)

- **Example**: load the banking.csv file available in Canvas 

```{r, eval=TRUE, echo=TRUE}
# Define the Filename 
filename <- "banking.csv"
# Read the data set into R
banking_data <- read.csv(file = filename, header=TRUE)
# Store it as a tibble data frame 
banking_df <- tibble(banking_data)
# Show it 
banking_df
```

# Exporting data from R 

- Once you have completed data cleaning or your analysis you might want to export your data set. 

- You can export your data as a csv, xlsx, rms, or any other supported file extensions. 

- CSV is recommended (widely used and light). 

- Suppose I want to export data only from debit transactions. 

- We can use the filter function from *dplyr*

```{r, eval=TRUE, echo=TRUE}
banking_debit <- filter(banking_df, Debit == 1)
```
- Now we can export the data using *write.csv*

```{r, eval=TRUE, echo=TRUE}
# Define the Filename 
name_output <- "banking_debit.csv"
# Write data to your working directory
write.csv(x=banking_debit, file = name_output)
```

- Verify in your folder, the data should be stored there. 

# Coding Exercise

- Best way to learn R is to practice writing code! 

- Perform all the tasks at the Lab Exercise. 

- You are going to use the `General Social Survey 2008.csv` file. 

# Appendix: Accesing Elements of Data Structures 

- You can retrieve only some elements of any data structures. 

```{r, eval=TRUE, echo=TRUE}
# Shows second element of vector1
vector1[2]
# Shows first 3 elements of vector1
vector1[1:3]
```

- Notation could differ slightly for lists. Note the double brackets. 

```{r, eval=TRUE, echo=TRUE}
# Shows first element of list1
list1[[1]]
```

- You can access full rows (or columns of a matrix)

```{r, eval=TRUE, echo=TRUE}
# Shows first row of the matrix stored in list1
list1[[2]][1,]
```
